---
title: "Mastering Colour Selection"
output: html_document
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

This document provides guidance on using the `colour_display()`, `colour_pal()`, and `colour_prep()` functions from an R package to create and visualise colour palettes effectively.
```{r setup, include = FALSE}
library(scgUtils)
```

Colour plays a crucial role in data visualisation. It not only makes plots more aesthetically pleasing but also aids in communicating data patterns and insights more effectively. The functions `colour_display()`, `colour_pal()`, and `colour_prep()` are designed to help users select and apply colour palettes with ease.
<br>
<br>

### `colour_display()`

The `colour_display()` function visualises colour options from a specified palette. It integrates with `colour_pal()` to fetch colour data and uses an internal helper function, `contrast_test()`, to ensure text readability on coloured bars.

```{r colour_display, warning=FALSE}
# Example: Visualise the 'polUK' palette
colour_display("polUK")
# Example: Visualise an individual colour
colour_display("Jaffa")
# Example: Visualise all individual colours
colour_display("All")
# Example: View a sequential colour palette with a continuous scale
colour_display("seqGreen", 7, type = "continuous")
```
<br>
<br>

### `colour_pal()`
The `colour_pal()` function generates colour palettes for graphs. It supports sequential, diverging, categorical, and political palettes, allowing users to retrieve individual colours or ranges of colours based on the specified palette.

```{r colour_pal}
# Example: Return full categorical extended palette
colour_pal("catExtended")

# Example: Return an individual colour
colour_pal("Jaffa")

# Example: Return a diverging palette with assigned categories
colour_pal("divBlueGreen", 5, c("Very Likely", "Likely", "Neutral", "Unlikely", "Very Unlikely"))
```
<br>
<br>

### `colour_prep()`
`colour_prep()` prepares a colour palette for visualising data by assigning colours to unique values across specified columns in a data frame. It is particularly useful for creating consistent colour mappings in visualisations.

```{r colour_prep}
# Example data frame
df <- data.frame(
  Category1 = c("Scottish National Party (SNP)", "Labour", "Conservative", "Other"),
  Category2 = c("An independent candidate", "Invalid_Party_Name", "Other", "Conservative")
)

# Prepare a colour palette for the unique values in Category1 and Category2
colour_palette <- colour_prep(df, c("Category1", "Category2"))

# Print the prepared colour palette
print(colour_palette)
```