---
title: "Other Functions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```
Survey data requires cleaning, iterating, and analysing. The following provides available functions to help with these processes.

### Grouped Frequency
The `grp_freq` function calculates the frequency of a variable in survey data by specified groups. This function is in place of [dplyr's](https://dplyr.tidyverse.org/reference/summarise.html) `group_by(y) %>% summarise(x = sum(x))` function.

```{r grp_freq}
# GROUPED FREQUENCY DATA
# Sum unweighted data by 2 groups
df <- grp_freq(dataset,
               group = c("age_categories","gender"))
#    age_categories gender Freq
# 1        18 to 24   Male   84
# 2        25 to 34   Male  155
# 3        35 to 44   Male  180
# 4        45 to 54   Male  115
# 5        55 to 54   Male   38
# 6             65+   Male    6
# 7        18 to 24 Female   83
# 8        25 to 34 Female  153
# 9        35 to 44 Female  129
# 10       45 to 54 Female   57
# 11       55 to 54 Female   24
# 12            65+ Female    6

# Sum weighted data by 1 group and change the column names
df <- grp_freq(dataset,
               group = c("age_categories"),
               weight = "wgtvar",
               set_names = c("Age","n"))
#         Age         n
# 1  18 to 24  86.29264
# 2  25 to 34 225.19210
# 3  35 to 44 275.61526
# 4  45 to 54 280.12479
# 5  55 to 54 100.97521
# 6       65+  61.80000
```
<br>

### Grouped Means
The `grp_mean` function calculates the mean of a variable in survey data by specified groups. For non-grouped mean data, use `mean(var)` or `weighted.mean(var,weight)`.

```{r grp_mean}
# GROUPED MEAN DATA
# Return a averages of a variable by group (weighted or unweighted)
df <- grp_mean(dataset,
               var = "age",
               group = "gender",
               weight = "wgtvar")
#   gender     Mean
# 1   Male 42.78670
# 2 Female 41.06441

# NB for non-grouped averages, use mean(var) or weighted.mean(var, weight)
```
<br>

### Grid-Style Survey Questions
The `grid_vars` function takes grid style survey questions and returns a data frame with the frequency and percentage of each response variable. This is ordinarily required when survey's ask questions such as "select all that apply". The function helps prepare data for analysis, comparison, and visualisation.
```{r griv_vars}
# Example data:
#     wgt gender  Q1a Q1b Q1c
# 1  0.61   Male  Yes  No Yes
# 2  0.22 Female  Yes Yes Yes
# 3  1.81   Male   No  No Yes
# 4  0.90   Male  Yes  No  No
# 5  1.63 Female   No Yes  No
# 6  1.00 Female   No  No Yes
# ...

# Create list containing the column name and its associated variable name
vars <- list(Q1a = "Art",
             Q1b = "Automobiles",
             Q1c = "Birdwatching")

# Create data frame with binary variables
grid_vars(dataset,
            vars = vars,
            group = "gender",
            weight = "wgtvar")
#      Question Response gender     Freq Perc
#           Art      Yes   Male 275.3617   46
#   Automobiles      Yes   Male 320.1372   53
#  Birdwatching      Yes   Male 310.4357   52
#           Art      Yes Female 204.7525   48
#   Automobiles      Yes Female 212.0209   49
#  Birdwatching      Yes Female 203.9380   47
# ...

# NB. To see as a plot, see "Binary Plot" below.
```