---
title: "Getting Started"
---

After installation, load `scgUtils` to access its functionalities:

```{r}
library(scgUtils)
```

### Key Functions
#### Data Processing

* [`get_file()`](https://sarahcgall.github.io/scgUtils/reference/get_file.html): Retrieve and preprocess files from various sources.

#### Data Exploration

* [`compile()`](https://sarahcgall.github.io/scgUtils/reference/compile.html): Combine multiple data frames for comprehensive analysis.
* [`grid_vars()`](https://sarahcgall.github.io/scgUtils/reference/grid_vars.html): Transform binary grid survey questions for deeper insights.

#### Visualisation

* [`plot_parliament()`](https://sarahcgall.github.io/scgUtils/reference/plot_parliament.html): Create parliament-style visual representations.
* [`plot_popn()`](https://sarahcgall.github.io/scgUtils/reference/plot_popn.html): Depict population profiles with customisability.

#### Styling

* [`colour_display()`](https://sarahcgall.github.io/scgUtils/reference/colour_display.html): Explore and display colour options for plots.
* [`colour_prep()`](https://sarahcgall.github.io/scgUtils/reference/colour_prep.html): Prepare named colour palettes for datasets.

### Example Usage
Here's a quick example of using `scgUtils` to analyse and visualise survey data:

```{r}
# Load sample data
df <- get_file("inst/extdata/survey.sav")

# Explore data with a crosstab
crosstab(data = df,
         rowVar = "p_eurefvote",
         colVar = "p_edlevel",
         weight = "wt",
         format = "df_wide",
         round_decimals = 2,
         statistics = TRUE,
         plot = TRUE
)
# [1] "p_eurefvote x p_edlevel: Chisq = 371.026 | DF = 10 | Cramer's V = 0.243 | p-value = 0"
#         p_eurefvote Undergraduate A-level Below GCSE  GCSE No qualifications Postgrad Total
# 1 I voted to remain         61.23   40.58      27.47 30.60             23.49    73.47 47.92
# 2  I voted to leave         38.25   59.12      72.53 69.13             76.51    26.42 51.78
# 3        Don't know          0.52    0.30       0.00  0.27              0.00     0.10  0.31
```
```{r example, include=FALSE}
knitr::opts_chunk$set(
  fig.path = "vignettes/articles/img/crosstab_example.png"
)
```
