---
title: "Using Survey Data"
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

This article aims to guide you through the process of uploading, processing, and exploring survey data.

```{r setup, include = FALSE}
library(scgUtils)
```
<br>

### Step 1: Upload
If you are using the sample dataset within this package, use the following:
```{r upload_data}
df <- scgElectionsNZ::get_data("survey")
```

<br>
Alternatively, if you are uploading your own data, you can use the `haven` package to read an SPSS file in R.

```{r upload_data_alternative, eval = FALSE}
df <- haven::read_sav("your_survey_data.sav")
```

*Troubleshooting `read_sav()`:*
NB if an error occurs which relates to encoding, try adding `encoding="latin1"`. Once you have uploaded the data, view which string is causing the error. This may have been caused by `NAs` which will require you to convert them from a string to `NA`.
```{r upload_data_troubleshooting, eval = FALSE}
df <- haven::read_sav("your_survey_data.sav", encoding="latin1")

# Additional step to amend error across entire dataset (all columns)
df[df == "__NA__"] <- NA
```
<br>

### Step 2: View
View the first 6 rows of the data file and the details of each column
```{r view_data2, eval = FALSE}
# NB first 3 variables shown only
head(df[,1:3])
# A tibble: 6 x 3
#      id    wt turnoutUKGeneral
#   <dbl> <dbl> <dbl+lbl>
# 1     7 0.376 5 [Very likely that I would vote]
# 2    14 0.553 5 [Very likely that I would vote]
# 3    15 0.712 5 [Very likely that I would vote]
# 4    18 0.440 4 [Fairly likely]
# 5    19 0.361 5 [Very likely that I would vote]
# 6    24 1.69  5 [Very likely that I would vote]

str(df)
# tibble [5,000 x 65] (S3: tbl_df/tbl/data.frame)
#  $ id                  : num [1:5000] 7 14 15 18 19 24 30 39 44 46 ...
#   ..- attr(*, "label")= chr "ID"
#   ..- attr(*, "format.spss")= chr "F8.0"
#  $ wt                  : num [1:5000] 0.376 0.553 0.712 0.44 0.361 ...
#   ..- attr(*, "label")= chr "New Weight Wave 25"
#   ..- attr(*, "format.spss")= chr "F12.2"
#  $ turnoutUKGeneral    : dbl+lbl [1:5000]    5,    5,    5,    4,    5,    5,
#  5,    5,    5,...
#    ..@ label      : chr "Likelihood to vote in general election"
#    ..@ format.spss: chr "F31.0"
#    ..@ labels     : Named num [1:6] 1 2 3 4 5 ...
#    .. ..- attr(*, "names")= chr [1:6] "Very unlikely that I would vote"
#    "Fairly unlikely" "Neither likely nor unlikely" "Fairly likely" ...
# ...
```
<br>

You can also view the original questionnaire and values easily by using the `sjPlot` package.
```{r view_data1, eval = FALSE}
# NB first 10 variables shown only
sjPlot::view_df(df[,1:10])
```
````{=html}
<iframe width="100%" height="850"
  src="img/questionnaire.html"
  frameborder="0"
  allowfullscreen></iframe>
````
<br>

### Step 3: Clean
To add the labels to the variables, use the `labelled` package.
```{r clean_data1}
df <- labelled::unlabelled(df)
```

The variables have now been labelled and converted into factors:
```{r clean_data2, eval = FALSE}
# NB first 3 variables shown only
head(df[,1:3])
# A tibble: 6 x 3
#      id    wt turnoutUKGeneral
#   <dbl> <dbl> <fct>
# 1     7 0.376 Very likely that I would vote
# 2    14 0.553 Very likely that I would vote
# 3    15 0.712 Very likely that I would vote
# 4    18 0.440 Fairly likely
# 5    19 0.361 Very likely that I would vote
# 6    24 1.69  Very likely that I would vote

# Check the class of each variable
lapply(df,class)
# $id
# [1] "numeric"
#
# $wt
# [1] "numeric"
#
# $turnoutUKGeneral
# [1] "factor"
# ...

# OR
str(df)
# tibble [5,000 x 65] (S3: tbl_df/tbl/data.frame)
#  $ id                  : num [1:5000] 7 14 15 18 19 24 30 39 44 46 ...
#   ..- attr(*, "label")= chr "ID"
#   ..- attr(*, "format.spss")= chr "F8.0"
#  $ wt                  : num [1:5000] 0.376 0.553 0.712 0.44 0.361 ...
#   ..- attr(*, "label")= chr "New Weight Wave 25"
#   ..- attr(*, "format.spss")= chr "F12.2"
#  $ turnoutUKGeneral    : Factor w/ 6 levels "Very unlikely that I would vote",..: 5 5 5 4 5 5 5 5 5 5 ...
#   ..- attr(*, "label")= chr "Likelihood to vote in general election"
# ...
```
<br>

### Step 4: Explore
To view the population profile of the respondents,
```{r popn_plot}
plot_popn(data=df,
          age_groups="ageGroup",
          gender="gender",
          weight="wt",
          age_int="age"
)
```
