---
title: "Using scgUtils with Survey Data"
---
```{=html}
<style>
  .content p {
    margin-top: 30px !important;
  }

  details {
    margin-bottom: 40px;
  }
</style>
```
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```
<br>
This article will take you through the steps of uploading, processing, and exploring survey data using the `scgUtils` package in R. Our journey will cover the nuances of handling survey datasets, from the initial loading of the data to the advanced stages of analysis and visualisation.
<br>

```{r setup, include = FALSE}
library(scgUtils)
library(magrittr)
```
### Step 1: Upload the Sample Dataset
The process begins with loading your dataset. The `scgUtils` package offers two main functions for this purpose: `get_data()` and `get_file()`.
<br>
<br>

**Using `get_data()`:** This function is ideal for loading datasets directly from R packages. It streamlines the process of importing and naming your dataset in the R environment.

```{r upload_data}
# Example of loading and preprocessing a dataset
df <- get_data("survey") %>%
  labelled::unlabelled() %>% # Convert 'haven_labelled' data to standard format
  process_factors() # Remove unused factor levels
```
*Note: The sample data is a subset of the British Electoral Survey. For full data, visit [British Election Study](https://www.britishelectionstudy.com/data-object/wave-25-of-the-2014-2023-british-election-study-internet-panel/).*
<br>
<br>
**Using `get_file()`:** When working with external data files, such as .sav or .csv, get_file() becomes invaluable. It not only imports the data but also preprocesses it, handling special characters and facilitates the conversion of specialised data types, such as `haven_labelled`, into standard R formats, making it a robust choice for various data sources.
```{r upload_data1, eval = FALSE}
# Using the  `get_file function option which includes the above preprocessing.
df <- get_file("inst/extdata/survey.sav")

head(df[, 1:6])
```
```{r upload_data_table, echo = FALSE}
knitr::kable(head(df[, 1:6]))
```

*For detailed uploading instructions, refer to the [Uploading & Cleaning Data article](https://sarahcgall.github.io/scgUtils/articles/uploading-data.html).*
<br>

---

### Step 2: Viewing the Full Dataset
After uploading the data, it's important to understand its structure and content. The `sjPlot` package's [`view_df()`](http://www.strengejacke.de/sjPlot/reference/view_df.html) function provides an interactive HTML view of your dataset, allowing for an immediate and comprehensive examination of the data's attributes, frequencies, and percentages. This step is crucial for identifying the nature of variables, understanding their distribution, and planning further data processing strategies.

```{r view_data1, eval = FALSE}
sjPlot::view_df(df[, 1:10],  # NB first 10 variables shown only in this example
                weight.by = "wt",
                show.type = TRUE, # show whether variable is numeric or categorical
                show.wtd.frq = TRUE, # display weighted frequency
                show.wtd.prc = TRUE, # display weighted %
)
```
````{=html}
<iframe width="100%" height="850"
  src="img/questionnaire.html"
  frameborder="0"
  allowfullscreen></iframe>
````
---

### Step 3: Weighting the Data
Survey data analysis sometimes necessitates weighting to address issues like sample design or response biases. Future developments in `scgUtils` will introduce user-friendly functions for applying weights, ensuring your analysis reflects the intended population accurately.

---

### Step 4: Processing the Data
Survey data often requires specific processing steps to ensure it is analysis-ready. This may involve creating new variables, recoding factors, handling missing values, and more. Future developments of the `scgUtils` package will aim to simplify these tasks, providing tools that streamline data preparation and enhance its quality for subsequent analysis.

---

### Step 5: Exploring the Data
Exploring survey data effectively demands a nuanced approach to both numeric and categorical data. The `scgUtils` package, complemented by base R functionalities, offers a comprehensive toolkit for this exploration.
<br>
<br>

#### **Numeric data**
Numeric data, such as age, income, or survey ratings, can reveal significant trends and patterns when analysed correctly.
<br>
<br>

***Summary Statistics***<br>
Begin with `summary()` for a quick overview, offering key statistical measures.
```{r numeric1}
summary(df$age)
```
<br>

For group-wise insights, `tapply()` combined with `summary()` allows you to dissect the data based on categories like gender or education level, providing a clearer understanding of distribution across different segments.
```{r numeric1_1}
# By group:
tapply(df$age, df$gender, summary)
```
<br>

***Mean Calculation***<br>
Average calculations, both weighted and unweighted, are crucial in survey analysis. Utilise `mean()` for simple averages and `weighted.mean()` for more complex scenarios where survey design needs to be accounted for.
```{r numeric2}
# Unweighted:
mean(df$age)

# Weighted:
weighted.mean(df$age, df$wt)
```
<br>

***Grouped Mean with `grp_mean()`***<br>
For advanced analysis, `grp_mean()` elegantly calculates group-wise means. It simplifies the process of aggregating data across one or more categorical variables, offering an efficient alternative to more verbose methods like those in `dplyr`.
```{r numeric3, eval = FALSE}
# By a single group:
grp_mean(df,
         meanVar = "age",
         groups = "gender",
         weight = "wt" # optional
)

# `dplyr` equivalent:
# df %>%
#   group_by(gender) %>%
#   summarise(Mean = weighted.mean(age, wt)) %>%
#   ungroup()
```
```{r numeric3_table, echo = FALSE}
knitr::kable(
  grp_mean(df,
           meanVar = "age",
           groups = "gender",
           weight = "wt"
  )
)
```
```{r numeric4, eval = FALSE}
# By many groups:
grp_mean(df,
         meanVar = "age",
         groups = c("gender", "partyId"),
         weight = "wt", # optional
         set_names = c("Gender", "Party Identification", "Average Age"), # optional: change names
         round_decimals = 2 # optional: round decimal places to 2 digits
) %>%
  head()

# `dplyr` equivalent:
# df %>%
#   group_by(gender, partyId) %>%
#   summarise(`Average Age` = weighted.mean(age, wt)) %>%
#   ungroup() %>%
#   rename(Gender = gender, `Party Identification` = partyId) %>%
#   head()
```
```{r numeric4_table, echo = FALSE}
knitr::kable(
  grp_mean(df,
           meanVar = "age",
           groups = c("gender", "partyId"),
           weight = "wt",
           set_names = c("Gender", "Party Identification", "Average Age"),
           round_decimals = 2
  ) %>% head()
)
```
<br>

#### **Categorical data**
Categorical data typically includes demographics or multiple-choice responses. Analysing these effectively unlocks insights into respondent behaviours and preferences.
<br>
<br>

***Grouped Frequencies with `grp_freq()`***<br>
`grp_freq()` shines in its ability to provide detailed frequency and percentage breakdowns across various groups. It's capable of handling both weighted and unweighted data, adding depth and precision to your categorical data analysis.
```{r categorical1, eval = FALSE}
# By a single group:
grp_freq(df,
         groups = "partyId",
         weight = "wt", # optional
         addPercent = TRUE # optional
) %>%
  head()

# `dplyr` equivalent:
# df %>%
#   group_by(partyId) %>%
#   summarise(Freq = sum(wt)) %>%
#   ungroup() %>%
#   mutate(Perc = Freq / sum(Freq)) %>%
#   head()
```
```{r categorical1_table, echo = FALSE}
knitr::kable(grp_freq(df,
                      groups = "partyId",
                      weight = "wt", # optional
                      addPercent = TRUE # optional
) %>% head())
```
```{r categorical2, eval = FALSE}
# By many group:
grp_freq(df,
         groups = c("partyId", "gender"),
         weight = "wt", # optional
         groupsPercent = "partyId", # optional
         round_decimals = 2 # optional: round decimal places to 2 digits
) %>%
  head()

# `dplyr` equivalent:
# df %>%
#   group_by(partyId, gender) %>%
#   summarise(Freq = sum(wt)) %>%
#   ungroup() %>%
#   group_by(partyId) %>%
#   mutate(Perc = Freq / sum(Freq)) %>%
#   ungroup() %>%
#   head()

```
```{r categorical2_table, echo = FALSE}
knitr::kable(grp_freq(df,
                      groups = c("partyId", "gender"),
                      weight = "wt", # optional
                      groupsPercent = "partyId", # optional
                      round_decimals = 2 # optional: round decimal places to 2 digits
) %>% head())
```
<br>

***Crosstabs***<br>
Crosstabulation is a fundamental technique in survey analysis, especially when examining relationships between categorical variables. For an overview of using crosstab functions within the `scgUtils` package, view the [Conducting Cross-Tabulation Analysis article](https://sarahcgall.github.io/scgUtils/articles/crosstab.html).
<br>
<br>

#### **Grid data**
Grid questions, common in surveys, pose unique analytical challenges due to their format.
<br>
<br>

***Handling Grid Data with `grid_vars()`***<br>
`grid_vars()` is tailor-made for such data, turning complex grid questions into analysable formats. It's particularly adept at handling "select all that apply" questions or grid-type responses, transforming them into a format conducive to comparison and visualisation with libraries such as [`ggplot2`](https://ggplot2.tidyverse.org/).
```{r grid1, eval = FALSE}
# Create a named list of the columns that relate to the question
vars <- list(likeSunak = "Rishi Sunak",
             likeStarmer = "Keir Starmer",
             likeCon = "Conservative Party",
             likeLab = "Labour Party",
             likeLD = "Lib Dems",
             likeSNP = "SNP",
             likePC = "Plaid Cymru",
             likeBrexitParty = "Brexit Party",
             likeGrn = "Green Party"
)

grid_vars(df,
          vars = vars,
          weight = "wt" # optional
) %>%
  head()
```
```{r grid1_table, echo = FALSE}
# Create a named list of the columns that relate to the question
vars <- list(likeSunak = "Rishi Sunak",
             likeStarmer = "Keir Starmer",
             likeCon = "Conservative Party",
             likeLab = "Labour Party",
             likeLD = "Lib Dems",
             likeSNP = "SNP",
             likePC = "Plaid Cymru",
             likeBrexitParty = "Brexit Party",
             likeGrn = "Green Party"
)

knitr::kable(grid_vars(df,
                       vars = vars,
                       weight = "wt" # optional
) %>% head()
)
```
<br>

***Implementing `grid_vars()` by Group***<br>
Enhance the functionality of `grid_vars()` by applying it with a group variable. This allows for dissecting responses across different demographic or categorical segments, providing richer, more targeted insights.
```{r grid2, eval = FALSE}
grid_vars(df,
          vars = vars,
          group = "gender", # optional
          weight = "wt" # optional
) %>%
  head()
```
```{r grid2_table, echo = FALSE}
knitr::kable(grid_vars(df,
                       vars = vars,
                       group = "gender",
                       weight = "wt"
) %>% head())
```
---

### Step 6: Visualising the Data
Effective visualisation is key in survey data analysis, offering a way to intuitively understand and communicate complex data patterns. The `scgUtils` package provides specialised functions like `plot_popn()` for demographic analysis and `plot_sankey()` for flow visualisation, helping you to not only understand your data but also to present it in a compelling and insightful manner.

For a broader spectrum of visualisation techniques and detailed guidance on effectively using colour in your plots, refer to the [Visualising Data article](https://sarahcgall.github.io/scgUtils/articles/visualising-data.html) and [Mastering Colour Selection article](https://sarahcgall.github.io/scgUtils/articles/creating-colour-palettes.html#colour_display). These resources provide additional insights into making the most of the `scgUtils` package for visualising complex survey data.
<br>

---

### Step 7: Presenting the Results
Presenting the results of your survey analysis in a clear and impactful way is crucial. The `scgUtils` package offers functionalities that aid in creating detailed and informative presentations.
<br>
<br>

#### **Tables**
Tables are fundamental tools for presenting complex data in a structured and easily interpretable format. View the [Conducting Cross-Tabulation Analysis article](https://sarahcgall.github.io/scgUtils/articles/crosstab.html) for more information on how to compile tables.
<br>
<br>

#### **PowerPoint Integration**
In future updates, `scgUtils` aims to incorporate capabilities for directly exporting analysis results into PowerPoint presentations. This functionality will facilitate seamless integration of your data findings into professional and engaging presentation formats, suitable for various audiences.
<br>
<br>

#### **Interactive Dashboards**
Another upcoming feature is the ability to create interactive dashboards directly from your survey data. Dashboards offer a dynamic way to explore and present data, allowing users to interact with the information, drill down into specifics, and gain a deeper understanding of the underlying patterns. This functionality will be a significant enhancement, providing a powerful tool for data storytelling and decision-making processes.