---
title: "Visualising Data"
output: html_document
---
```{=html}
<style>
  .content p {
    margin-top: 30px !important;
  }

  details {
    margin-bottom: 40px;
  }
</style>
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```
<br>
This article demonstrates the powerful visualisation capabilities of the `scgUtils` package, offering tools for diverse data presentations ranging from personality profiles to demographic and flow analyses.
<br>

```{r setup, include = FALSE}
library(scgUtils)
library(magrittr)
library(dplyr)
library(tidyr)

survey_df <- get_data("survey") %>%
  labelled::unlabelled() %>%
  process_factors()
```
### Sankey
Flow visualisation helps in understanding how different categories of respondents transition between various stages or choices. The `plot_sankey()` function is instrumental in depicting the flow of data, especially useful in understanding voting patterns or changes in preferences over time.
<br>
<br>

##### Preparing Data with `grp_freq()`
Before visualising, prepare your data using `grp_freq()`, which aggregates frequencies necessary for the Sankey diagram.
```{r flow1, eval = FALSE}
# Subset the required columns from the dataset
sankey_df <- survey_df[, c("wt", "generalElectionVote", "p_past_vote_2019")]

# Get the frequency
sankey_df <- grp_freq(sankey_df,
                      groups = c("generalElectionVote", "p_past_vote_2019"),
                      weight = "wt", # optional
                      round_decimals = 0, # optional
)
head(sankey_df)
# NB. The `dplyr` equivalent is:
# df %>%
#   group_by(generalElectionVote, p_past_vote_2019) %>%
#   summarise(Freq = sum(wt))
```
```{r flow1_table, echo = FALSE}
sankey_df <- survey_df[, c("wt", "generalElectionVote", "p_past_vote_2019")]
sankey_df <- grp_freq(sankey_df,
                      groups = c("generalElectionVote", "p_past_vote_2019"),
                      weight = "wt",
                      round_decimals = 0,
)
knitr::kable(head(sankey_df))
```
<br>

##### Customising the Sankey Diagram
The `plot_sankey()` function offers extensive customisation, allowing the diagram to be tailored to specific data narratives. The `colour_prep()` function enhances this customisation by facilitating the assignment of meaningful colours based on categories like political party affiliations. Such customisation not only improves the aesthetic appeal of the Sankey diagram but also boosts its interpretability and effectiveness in conveying complex data flows.
```{r flow2, eval = FALSE}
plot_sankey(sankey_df,
            source = "p_past_vote_2019", # on the left side
            target = "generalElectionVote", # on the right side
            value = "Freq",
            units = " votes",
            colours = colour_prep(df, c("generalElectionVote", "p_past_vote_2019"), pal_name = "polUK"),
            fontSize = 16, # change font size
            fontFamily = "Calibri", # default
            nodeWidth = 20, # default
            nodePadding = 10, # default
            margin = list(top = 0, right = 130, bottom = 0, left = 0), # adjust the margin
            width = 1200, # default
            height = 800, # default
            shiftLabel = NULL, # default
            heading = "Flow of Votes",
            sourceTitle = "2019 Vote",
            targetTitle = "VI"
) # %>%
# save from viewer to html
# htmlwidgets::saveWidget(file = "sankey_VI.html", selfcontained = TRUE)
```
````{=html}
<iframe width="100%" height="850"
  src="img/sankey_VI.html"
  frameborder="0"
  allowfullscreen></iframe>
````
<br>

### Parliament
Understanding the distribution of parliamentary seats among political parties is crucial for grasping the political landscape. The `plot_parliament()` function in `scgUtils` is designed to visualise this distribution in a semicircular parliament layout. It is particularly useful for illustrating the composition of a parliament following an election.
<br>
<br>

##### Basic Parliament
The basic usage of `plot_parliament()` involves creating a plot that shows the number of seats each party holds. This representation helps in quickly understanding the strength of each party within the parliament.
```{r parliament1, dpi=300, out.width = '100%', fig.height = 4}
# Prepare Data
de_parliament <- data.frame(
  Party = c("SPD", "Greens", "FDP", "The Left", "Other", "AfD", "CDU/CSU"),
  Result = c(206, 118, 92, 39, 1, 83, 97)
)

# Plot
plot_parliament(de_parliament,
                partyCol = "Party",
                seatCol = "Result",
                colours = c("#e3000f", "#409a3c", "#ffed00", "#be3075", "#dcdcdc", "#00a2de", "black") # optional
)
```
<br>

##### Adding a Percentage Bar
For a more detailed analysis, `plot_parliament()` can also include a percentage bar that shows the popular vote won by each party. This feature provides additional context to the seat distribution, reflecting how party popularity translates into parliamentary seats.
```{r parliament2, dpi=300, out.width = '100%', fig.height = 5}
# Prepare Data
uk_parliament <- data.frame(
  Party = c("Labour", "SNP", "Other", "Liberal Democrat", "Conservative"),
  Seats = c(202, 48, 24, 11, 365),
  Percentage = c(32.1, 3.9, 8.8, 11.6, 43.6)
)

# Plot
plot_parliament(uk_parliament,
                "Party",
                "Seats",
                "Percentage",
                majorityLine = TRUE, # add line down centre
                title = "2019 UK General Election", # add title
                subtitle = "Results", # add subtitle
                legend = "bottom", # add legend to bottom
                colours = colour_prep(uk_parliament, "Party", "polUK"), # match colours using `colour_prep()`
)
```
<br>
<br>
This plot offers an intuitive way to analyse election results, party strengths, and their representation in the parliament. The inclusion of a majority line further enhances the plot by delineating the threshold needed for a majority.
<br>
<br>

### Population
Understanding demographic distribution is vital in survey analysis. `plot_popn()` creates visual representations of population profiles.

##### Using `plot_popn`
The `plot_popn()` function is designed to visualise the population structure of your survey respondents. It creates a population pyramid showing distributions across gender and age groups. If a variable like average age (`meanVar`) is specified, the plot can also display this information, adding another layer of insight into the demographic composition.

```{r popn_plot, dpi=300, out.width = '100%', fig.height = 5}
plot_popn(data = survey_df,
          xVar = "gender",
          yVar = "ageGroup",
          weight = "wt", # optional
          meanVar = "age", # optional (must be numeric)
          addLabels = TRUE # to add % labels
)
```
<br>

##### Faceting by Group
Enhance your population pyramid by faceting the `plot_popn` plot by a specific group, such as voter turnout. This feature overlays the selected group's data onto the total population structure, providing a comparative view that highlights differences or similarities within subgroups.

```{r popn_plot2, dpi=300, out.width = '100%', fig.height = 7}
plot_popn(data = survey_df,
          xVar = "gender",
          yVar = "ageGroup",
          group = "turnoutUKGeneral",
          weight = "wt", # optional
          addLabels = TRUE # to add % labels
)
```
<br>

### Personality
The `plot_bigfive()` function returns a [ggplot2](https://ggplot2.tidyverse.org/reference/theme.html) chart to help visualise the personality profile of the survey data. This radar chart is primarily to visualise the Big Five personality traits (neuroticism, extroversion, openness, agreeableness, and conscientiousness) but can be amended for other quantitative data types with a scale between 0 and 100.

```{r dummydata_personality, echo = FALSE}
# Create dummy dataset
bigfive_df <- data.frame(Gender = c("Male", "Female", "Male", "Male", "Female", "Female", "Male", "Female"),
                         Weight = c(0.6, 0.8, 0.9, 1.0, 1.3, 1.7, 1.0, 0.99),
                         Neuroticism = c(60, 40, 30, 80, 20, 25, 50, 10),
                         Extroversion = c(75, 20, 35, 45, 50, 10, 60, 90),
                         Openness = c(50, 50, 45, 30, 65, 80, 10, 55),
                         Agreeableness = c(90, 30, 50, 20, 10, 75, 65, 35),
                         Conscientiousness = c(45, 50, 90, 10, 25, 30, 80, 40))
```

```{r bigfive_plot1, dpi=300, out.width = '100%', fig.height = 5}
# Create single plot using unweighted data
plot_bigfive(bigfive_df,
             bigfive = c("Neuroticism", "Extroversion", "Openness", "Agreeableness", "Conscientiousness"))
```
<br>

When a group is provided, the function returns faceted plots with the variables within the group plotted on top of the average. This provides an easy comparison between the variable and the rest of the cohort in the survey.

```{r plot_bigfive2, dpi=300, out.width = '100%', fig.height = 5}
# Create faceted plot using age groups and weighted data
plot_bigfive(bigfive_df,
             bigfive = c("Neuroticism", "Extroversion", "Openness", "Agreeableness", "Conscientiousness"),
             group = "Gender",
             weight = "Weight")
```
<br>

### Binary
The `plot_binary()` function visualises binary survey responses (e.g., "Yes" vs "No"). It is particularly effective for comparative analysis. This function utilises the `grid_vars()` function to help transform the data into the correct format.
```{r dummydata_binary, echo = FALSE}
set.seed(0)

# Initialize the data frame with binary responses
binary_df <- as_tibble(replicate(10, sample(c("Yes", "No"), 1000, replace = TRUE, prob = c(runif(1), 1-runif(1)))))

# Name the columns accordingly
names(binary_df) <- paste0("Q1", letters[1:10])

# Add a "Weight" column with values between 0.5 and 2.0
binary_df$Weight <- stats::runif(1000, min = 0.5, max = 2.0)

# Add a "Gender" column with approximately equal "Male" and "Female" values
binary_df$Gender <- factor(sample(c("Male", "Female"), 1000, replace = TRUE, prob = c(48,52)),
                           levels = c("Male", "Female"))
```
```{r plot_binary1, dpi=300, out.width = '100%', fig.height = 5}
# Create list for dummy data
vars <- list(Q1a = "Art",
             Q1b = "Automobiles",
             Q1c = "Birdwatching",
             Q1d = "Music",
             Q1e = "Reading",
             Q1f = "Cooking",
             Q1g = "Hiking",
             Q1h = "Watching Sport",
             Q1i = "Computers",
             Q1j = "Gaming"
)
# Create plot of total dataset using unweighted data
plot_binary(binary_df,
            vars = vars,
            value = "Yes"
)
```
<br>

When a group is provided, the function return faceted plots with the variables within the group plotted against the total. This provides an easy comparison between the variable and the rest of the cohort in the survey.
```{r plot_binary2, dpi=300, out.width = '100%', fig.height = 5}
# Create faceted plot using Gender and weighted data
plot_binary(binary_df,
            vars = vars,
            value = "Yes",
            group = "Gender", # optional
            weight = "Weight", # optional
            title = "Hobbies",
            subtitle = "by Gender"
)
```
<br>

### Waffle
Waffle plots provide a unique and compelling way to visualise categorical data, making them ideal for representing proportions or percentages in datasets, including survey responses. The `plot_waffle()` function in the `scgUtils` package offers a straightforward method to create these plots.
<br>
<br>
##### Basic Usage
The basic use of `plot_waffle()` involves creating a plot that represents the distribution of different categories. This method is especially effective for visually demonstrating the relative sizes of groups within a population.
```{r waffle1, dpi=300, out.width = '100%', fig.height = 4}
# Prepare Data
waffle_df <- data.frame(
    Category = c("A", "B", "C"),
    Count = c(30, 40, 30)
  )

# Plot
plot_waffle(waffle_df,
            group = "Category",
            values = "Count",
            isolateVar = "A" # show a single plot only
)
```
<br>

##### Using `plot_waffle()` with Survey Data
`plot_waffle()` is particularly adept at handling survey data, supporting both weighted and unweighted analysis. The function can automatically extract and display relevant labels from survey questions, enhancing the plot's interpretability.
```{r waffle2, dpi=300, out.width = '100%', fig.height = 6}
# Waffle plot with unweighted survey data
plot_waffle(survey_df %>% filter(p_socgrade != "Unknown"), # removing unknowns
            group = "p_socgrade",
            title = "p_socgrade"
)
```
<br>
<br>
For a more refined visual presentation, `plot_waffle()` allows customisation of colours and plot order. This flexibility is invaluable for aligning the visual aesthetic with specific data narratives or brand guidelines.
<br>
<br>

### Likert Plots
Likert scale responses will be visualised with upcoming features, providing nuanced insights into survey data. This function will contain four options for visualisations, including:

* `100% Stacked Bars`
* `Diverging with Neutrals`
* `Diverging without Neutrals`
* `Faceted Bars`

### Other Plots
Future updates to `scgUtils` will introduce other plots such as:

* `plot_dumbbell()` which can be used to compare two categories and view the differences between numeric data.
* `plot_wordcloud()` to highlight keywords from either qualitative or quantitative results.
* `plot_donut()` to illustrate numerical proportions.
* `plot_radar()` which will expand the capabilities of `plot_bigfive()` in order to allow the comparison of any numeric multivariate data.
* `plot_mekko()`to represent categorical data with multiple subcategories.

